---
title: "MESS"
author: "Suzie Brown and Marco Palma"
date: "26 October 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data wrangling
```{r}
### http://www.sciencedirect.com/science/article/pii/S0140673605666949?via%3Dihub LANCET ARTICLE MARSON 2005

mess <- read.csv("MESSdat.csv", sep="\t", header=T)
#summary(mess)
#head(mess)
mess$trt<-as.factor(mess$trt)
mess$sex<-as.factor(mess$sex)
```

We see that there are many NAs in the columns relating to dates, but this is not shown on the summary because they are formatted as character. We first format the dates as dates.
```{r}
mess$d1sp<-as.Date(mess$d1sp,"%d/%m/%y")
mess$d1cp<-as.Date(mess$d1cp,"%d/%m/%y")
mess$d1ps<-as.Date(mess$d1ps,"%d/%m/%y")
mess$d1myo<-as.Date(mess$d1myo,"%d/%m/%y")
mess$d1ab<-as.Date(mess$d1ab,"%d/%m/%y")
mess$d1aab<-as.Date(mess$d1aab,"%d/%m/%y")
mess$d1tc<-as.Date(mess$d1tc,"%d/%m/%y")
mess$d1oth<-as.Date(mess$d1oth,"%d/%m/%y")
mess$d1seiz<-as.Date(mess$d1seiz,"%d/%m/%y")
mess$rand<-as.Date(mess$rand,"%d/%m/%y")
```

Next we want to see which way round the censoring indicator is used. We see that where the censoring indicator `ind1yr` is 1, the time to one year remission `int1yr` has minimum 365, whereas where `ind1yr` is 0, the minimum is 1. From this we deduce that the indicator is 1 when the variable is observed and 0 if the variable is censored.
```{r}
summary(mess$int1yr[mess$ind1yr==1])
summary(mess$int1yr[mess$ind1yr==0])
```

Plotting the age at randomisation `ager` we see that it is positively skewed. This is plausible since it is a study of people with single seizure or recent diagnosis of epilepsy, and it is more likely that an individual has their first seizure at a younger age.
```{r}
hist(mess$ager,breaks=20)
```

We see that the number of subjects having each treatment is roughly equal, as was the design of the study.
```{r}
table(mess$trt)
```

We assume this variable must be an identifying number for the clinic where the subject entered the trial. As such, we format it as a factor.
```{r}
table(mess$centre)
mess$centre <- as.factor(mess$centre)
# maybe include in the model, by grouping them together? but it's a bit arbitrary - would be nice to know if the numbers are related to geographical location.
```

We verify that the same data are missing from `d1seiz` and `period`. These were probably subjects who couldn't remember the date of their first seizure, and/or whose medical records were missing. This suggests they are missing not at random - for instance if the subject can't remember the date it is likely to be a long time ago, i.e. higher values of `period`.
```{r}
all.equal(which(is.na(mess$d1seiz)),which(is.na(mess$period)))
```

Next we find that the binary results for various medical tests have been stored as 1/2 = yes/no, so we convert these to 0/1 = no/yes.
```{r}
mess$eeg <- as.factor(ifelse(mess$eeg==2,0,mess$eeg))
mess$abeeg <- as.factor(ifelse(mess$abeeg==2,0,mess$abeeg))
mess$nsab <- as.factor(ifelse(mess$nsab==2,0,mess$nsab))
mess$gparsp <- as.factor(ifelse(mess$gparsp==2,0,mess$gparsp))
mess$gparnsp <- as.factor(ifelse(mess$gparnsp==2,0,mess$gparnsp))
mess$fparsp <- as.factor(ifelse(mess$fparsp==2,0,mess$fparsp))
mess$fparnsp <- as.factor(ifelse(mess$fparnsp==2,0,mess$fparnsp))
mess$swab <- as.factor(ifelse(mess$swab==2,0,mess$swab))
```

There is a mystery as to how some patients who have not had an EEG have recorded an abnormal EEG result. There are several possible explanations, but we will assume the patients with an abnormal EEG have had an EEG. We adjust the `eeg` varaible accordingly. Since there are only five subjects in this category it shoudln't substantially affect any results.
```{r}
t.eeg <- table(mess$eeg,mess$abeeg)
dimnames(t.eeg) <- list(EEG=c(0,1), abEEG=c(0,1))
t.eeg

mess$eeg[mess$eeg==0 & mess$abeeg==1] <- 1
```

Split age into bins.
```{r}
mess$agecat <- cut(mess$ager,c(-1,5,9,19,29,39,49,59,69,100))
```

For now, these are the covariates we have decided to work with. We are dropping all of the date variables because we suspect the absolute time does not matter, and we have relative times e.g. `period` saved in other variables. We have also dropped the sub-categories for different types of seizures, retaining just the number of tonic-clonic seizures and the overall number of seizures. We also drop the sub-categories of EEG abnormality, retaining just the overall indicator. We do not include the centre, because there are too many levels to find out anything useful since we don't know whether they are related geographically, for instance.
```{r}
#covariates<-mess[,c(12:14,32,34)]  ### INCLUDE ,37 when you understand the answer to the previous question 
```


# Check balance of randomisation
```{r}
par(mfrow=c(1,2))

t.sex <- table(mess$sex,mess$trt)
colnames(t.sex)<-c('immediate','deferred')
rownames(t.sex)<-c('male','female')
plot(t.sex, main="randomisation by sex", ylab="treatment")
mtext("sex",3)
text(0.25,0.75,t.sex[1,1])
text(0.75,0.75,t.sex[2,1])
text(0.25,0.25,t.sex[1,2])
text(0.75,0.25,t.sex[2,2])

t.age <- table(mess$agecat, mess$trt)
colnames(t.age)<-c('immediate','deferred')
plot(t.age, main="randomisation by age", ylab="treatment")
mtext("age",3)
```


# Exploratory - Kaplan-Meier curves

We plot Kaplan-Meier curves of the first seizure after randomization, grouped by treatment and sex. The plot show that the survival probability for those who received a deferred treatment is lower than for those immediately treated, at least for the first 5 years after the randomization. For what concerns the sex of the patients, it seems not to have a significant effect on the survival probability.

```{r}
library(survival)
mess$surv1 <- with(mess, Surv(intseiz1, indseiz1 == 1))
mfit <- survfit(surv1 ~ 1, conf.type = "log-log", data = mess)
#options(survfit.print.mean =T)
#plot(mfit, mark.time=T, xlab = "Days since randomization",ylab = "Survival")
#summary(mfit,censored=T)
par(mfrow=c(1,2))
mfit.trt <- survfit(surv1 ~ trt, conf.type = "log-log", data = mess)
plot(mfit.trt, conf.int = T, mark.time=F, col = 2:3, xlab = "Days since randomization", ylab = "Survival")
legend("topright", legend = c("Immediate","Deferred"), col = 2:3, lty=1)
mfit.sex <- survfit(surv1 ~ sex, conf.type = "log-log", data = mess)
plot(mfit.sex, conf.int = T, mark.time=F, col = c("orange","brown"), xlab = "Days since randomization",ylab = "Survival")
legend("topright", legend = c("Male","Female"), col = c("orange","brown"), lty=1)

#lfit <- survreg(surv1 ~ trt, data=mess)
#pct <- 1:98/100   # The 100th percentile of predicted survival is at +infinity
#ptime <- predict(lfit, newdata=mess, type='quantile',p=pct, se=TRUE)
#matplot(ptime$fit, 1-pct,xlab="Months", ylab="Survival", type='l', lty=c(1,2,2), col=1)
```

# Impute missing values ??

# I'm just exploring which covariates we might like to include...
```{r}
# par(mfrow=c(3,2))
# plot(survfit(Surv(mess$intseiz1,mess$indseiz1, type="right") ~ mess$trt), main="treatment")
# plot(survfit(Surv(mess$intseiz1,mess$indseiz1, type="right") ~ cut(mess$ager, c(-1, 20,40,60,80,100))), main="age", conf.int=T, col=rep(1:10, each=3))
# plot(survfit(Surv(mess$intseiz1,mess$indseiz1, type="right") ~ mess$sex), main="sex")
# plot(survfit(Surv(mess$intseiz1,mess$indseiz1, type="right") ~ mess$eeg), main="EEG")
# plot(survfit(Surv(mess$intseiz1,mess$indseiz1, type="right") ~ mess$abeeg), main="abnormal EEG")
```
We see that sex makes no difference, nor does having/not having an EEG. But as we might expect, the treatment has an effect on outcome, as does an abnormal EEG result. It is hard to see whether age makes a difference but perhaps it does.

# Transformations of some variables
We plotted histograms of each (non-binary) variable we were considering including. The variables `nseiz` and `period` are highly positively-skewed. This is not surprising since it is a study on early epilepsy, so most subjects have only had one seizure and the frist seizure was not long before entering the study.
```{r}
par(mfrow=c(2,2))
hist(mess$intseiz1, breaks=20)
hist(mess$ager, breaks=20)
hist(mess$nseiz, breaks=20) #this one should be transformed
hist(mess$period, breaks=20) #and this one
```
We will try transforming `period` and `nseiz`.
```{r}
#par(mfrow=c(1,2))
mess$period.tr <- log(mess$period + 1)
mess$nseiz.tr <- mess$nseiz > 1
hist(mess$period.tr, breaks = 20)
#hist(mess$nseiz.tr)
```

# Cox proportional hazards model

First we fit a naive proportional hazards model with the covariates of interest. We see that the treatment `trt` is significant, as is the number `nseiz` of seizures prior to entering the study. This is to be expected since subjects with a tendency for more frequent seizures are likely to go less time before the first post-randomisation seizure regardless of treatment. The elapsed time between the subject's first seizure and their entry into the trial is slightly significant, which is also unsurprising.
```{r}
mycox <- coxph(Surv(mess$intseiz1,mess$indseiz1,type="right") ~ trt + ager + sex + nseiz + period, data=mess)
summary(mycox)
```
Next we assess whether proportional hazards is a suitable assumption for these data. We see from the residual plots that the treatment seems to have a decreasing effect over time, but the residuals of the other covariates don't seem to vary over time. This is confirmed by the small p-value for `trt` in the test of the proportional hazards assumption. The treatment effect decreasing over time is indicative that a proportional hazards model is not suitable, but we will try a few fixes now.
```{r}
cox.zph(mycox)
par(mfrow=c(3,2))
plot(cox.zph(mycox))
```

We try adding into the model an interaction term of treatment with time. We see, as expected, that the interaction term is very significant. However the proportional hazards model is still rejected, so even with the interaction term we see that the Cox proportional hazards model is not suitable.
```{r}
mycox2 <- coxph(Surv(mess$intseiz1,mess$indseiz1,type="right") ~ trt + ager + sex + nseiz + period + trt:intseiz1, data=mess)
summary(mycox2)
cox.zph(mycox2)
```

Next we will fit the model using the transformed versions of `period` and `nseiz`. The significant covariates are pretty much the same as in the first model, however the assumption of proportional hazards is not supported in the `trt` and `nseiz` terms.
```{r}
mycox3 <- coxph(Surv(mess$intseiz1,mess$indseiz1,type="right") ~ trt + ager + sex + nseiz.tr + period.tr, data=mess)
summary(mycox3)
cox.zph(mycox3)
```

Finally, we will include an interaction term between `trt` and `nseiz`, since it is plausible that the treatment may have a different effect on people who suffer more severely (indicated by having had more seizures). With this model the proportional hazards assumption is satisfied. However, including the interaction term has reduced the treatment effect, and indeed the interaction term is only slightly significant.
```{r}
mycox4 <- coxph(Surv(mess$intseiz1,mess$indseiz1,type="right") ~ trt + ager + sex + nseiz.tr + period.tr + trt:nseiz.tr, data=mess)
summary(mycox4)
cox.zph(mycox4)
```


# Aalen's additive model

```{r}
library(timereg)
modaalen<-aalen(surv1 ~ trt + const(ager) + const(sex) + const(nseiz) + const(period) + const(abeeg), data = mess , n.sim=100)
summary(modaalen)
#par(mfrow=c(3,3))
plot(modaalen)
#modaalen2<-aalen(surv1 ~ trt + const(ager) + const(sex) + nseiz + const(period) + const(abeeg), data = mess , n.sim=100)
#plot(modaalen2)
```


# Parametric survival models

```{r}
modpar1 <- survreg(surv1 ~ trt + ager + sex + nseiz + period + abeeg, mess, dist="exponential")
#summary(modpar1)
modpar2 <- survreg(surv1 ~ trt + ager + sex + nseiz + period + abeeg, mess, dist="weibull")
#summary(modpar2)
modpar3 <- survreg(surv1 ~ trt + ager + sex + nseiz + period + abeeg, mess, dist="lognormal")
#summary(modpar3)
modpar4 <- survreg(surv1 ~ trt + ager + sex + nseiz + period + abeeg, mess, dist="loglogistic")
#summary(modpar4)

# library(flexsurv)
# modpar1 <- flexsurvreg(surv1 ~ trt + ager + sex + nseiz + period, data = mess, dist="weibull")
# summary(modpar1)
# modpar2 <- flexsurvreg(surv1 ~ trt + ager + sex + nseiz + period, data = mess, dist="lnorm")
# summary(modpar2)
# modpar3 <- flexsurvreg(surv1 ~ trt + ager + sex + nseiz + period, data = mess, dist="llogis")
# summary(modpar3)
# modpar5 <- flexsurvreg(surv1 ~ trt + ager + sex + nseiz + period, data = mess, dist="gengamma")
# summary(modpar5)

parmodels<-matrix(NA,nrow=4,ncol=4)
parmodels[1,] <- c(summary(modpar1)$n, modpar1$loglik[1],summary(modpar1)$df, AIC(modpar1))
parmodels[2,] <- c(summary(modpar2)$n, modpar2$loglik[1],summary(modpar2)$df, AIC(modpar2))
parmodels[3,] <- c(summary(modpar3)$n, modpar3$loglik[1],summary(modpar3)$df, AIC(modpar3))
parmodels[4,] <- c(summary(modpar4)$n, modpar4$loglik[1],summary(modpar4)$df, AIC(modpar4))
colnames(parmodels) <- c("N.obs","Loglik","df","AIC")
rownames(parmodels) <- c("exponential","weibull","lognormal","loglogistic")
parmodels
```

Among the accelerated failure time (AFT) models, lognormal model is preferred.

```{r}
summary(modpar3)

# pct <- 1:98/100   # The 100th percentile of predicted survival is at +infinity
# ptime <- predict(modpar2, newdata=data.frame(trt=0,ager=24,sex=0,nseiz=1,period= 438.4), type='quantile',p=pct, se=TRUE)
# matplot(cbind(ptime$fit, ptime$fit + 2*ptime$se.fit, ptime$fit - 2*ptime$se.fit)/30.5, 1-pct,xlab="Months", ylab="Survival", type='l', lty=c(1,2,2), col=1)

dfNew <- data.frame(sex=factor(c(0,0)), trt=factor(c(0,1)),ager=24,nseiz=1,period = 438.4, abeeg = factor(c(1,1)))
percs <- (1:99)/100
FLogN <- predict(modpar3, newdata=dfNew, type="quantile", p=percs, se=TRUE)
matplot(cbind(FLogN$fit[1, ],
              FLogN$fit[1, ] - 2*FLogN$se.fit[1, ],
              FLogN$fit[1, ] + 2*FLogN$se.fit[1, ]), 1-percs,
        type="l", main=expression(paste("LogNormal-Fit ", hat(S)(t), " with SE")),xlim = c(0,3000),
        xlab="t", ylab="Survival", lty=c(1, 2, 2), lwd=2, col="blue")
matlines(cbind(FLogN$fit[2, ],
               FLogN$fit[2, ] - 2*FLogN$se.fit[2, ],
               FLogN$fit[2, ] + 2*FLogN$se.fit[2, ])/30.5, 1-percs, col="red", lwd=2)
legend(x="topright", lwd=2, lty=c(1, 2, 1, 2), col=c("blue", "blue", "red", "red"),
       legend=c("Trt 0 (immediate)", "+- 2*SE", "Trt 0 (deferred)", "+- 2*SE"))
```

Now let us consider the transformed variables.

```{r}
modpar1.tr <- survreg(surv1 ~ trt + scale(ager,scale = F) + sex + nseiz.tr + period.tr + abeeg, mess, dist="exponential")
modpar2.tr <- survreg(surv1 ~ trt + scale(ager,scale = F) + sex + nseiz.tr + period.tr + abeeg, mess, dist="weibull")
modpar3.tr <- survreg(surv1 ~ trt + scale(ager,scale = F) + sex + nseiz.tr + period.tr + abeeg, mess, dist="lognormal")
modpar4.tr <- survreg(surv1 ~ trt + scale(ager,scale = F) + sex + nseiz.tr + period.tr + abeeg, mess, dist="loglogistic")

parmodels.tr<-matrix(NA,nrow=4,ncol=4)
parmodels.tr[1,] <- c(summary(modpar1.tr)$n, modpar1.tr$loglik[1],summary(modpar1.tr)$df, AIC(modpar1.tr))
parmodels.tr[2,] <- c(summary(modpar2.tr)$n, modpar2.tr$loglik[1],summary(modpar2.tr)$df, AIC(modpar2.tr))
parmodels.tr[3,] <- c(summary(modpar3.tr)$n, modpar3.tr$loglik[1],summary(modpar3.tr)$df, AIC(modpar3.tr))
parmodels.tr[4,] <- c(summary(modpar4.tr)$n, modpar4.tr$loglik[1],summary(modpar4.tr)$df, AIC(modpar4.tr))
colnames(parmodels.tr) <- c("N.obs","Loglik","df","AIC")
rownames(parmodels.tr) <- c("exponential","weibull","lognormal","loglogistic")  
parmodels.tr #the difference to be interpreted in terms of chisq(1) = 3.84 
```

# Random survival forest

```{r}
library(randomForestSRC)
library(ggRandomForests)
rsf.surv1<-rfsrc(Surv(intseiz1, indseiz1) ~ trt + ager + sex + nseiz.tr + period.tr + abeeg, data = mess, ntree = 100, importance=TRUE,  tree.err = T)
print(rsf.surv1)
plot(rsf.surv1)
gg_dta <- gg_vimp(rsf.surv1)
plot(gg_dta)
```



